---
title: 'BHR Example: Burden Heritability in the BipEx dataset'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
## Description 

This file contains the code to run a BHR analysis of bipolar disorder, from publicly available summary statistics to final estimates.

Because this trait is binary, and we have only summary statistics, we compute per-allele effect sizes from case and control allele counts (Equation 33 in BHR paper). In the case of a continuous trait, one should compute per-allele effect sizes using a linear model.

## Loading and filtering variants.

The source file, "Bipex_variants_results.tsv", can be downloaded at this URL:
https://bipex.broadinstitute.org/downloads

The baseline model file, "baseline_model.txt", can be downloaded at this URL:
https://github.com/ajaynadig/bhr/tree/master/reference_files

```{r load}
baseline_model <- read.table("~/Documents/oconnor_rotation/rarevariantproject/final_manuscript_repo/bhr/reference_files/ms_baseline_oe5.txt")

bp_variantlevel_bipex <- bigreadr::fread2("~/Downloads/BipEx_variant_results.tsv.gz")

variant_filter = bp_variantlevel_bipex$group == "Bipolar Disorder" & #Filter to Bipolar Disorder counts
  bp_variantlevel_bipex$in_analysis == TRUE & #Use variant filter from Palmer et al, 2022
  !is.na(bp_variantlevel_bipex$gene_id) & #Remove variants with NA gene ID 
  str_detect(bp_variantlevel_bipex$locus, "^chr\\d") #Subset to autosomal variants

  
#Subset variants 
bp_variantlevel_bipex <- bp_variantlevel_bipex[variant_filter,
                                               c("gene_id",
                                                 "consequence",
                                                 "ac_case",
                                                 "ac_ctrl",
                                                 "locus",
                                                 "mpc")]  

```

## Wrangling summary statistics into format for BHR

```{r wrangle}
wrangle_sumstats <- function(table,n_cases,n_controls, var_filter) {
  
  #Filter to variants of interest
  table = table[var_filter,] 
  
  #Compute sample prevalence, will be used to compute per-sd beta
  prevalence = n_cases/(n_cases+n_controls) 
  
  #Compute variant MAF in cases, will be used to compute per-sd beta
  table$AF_case = table$ac_case/(2*n_cases) 
  
  #Compute variant MAFoverall, will be used to compute per-sd beta
  table$AF = (table$ac_case + table$ac_ctrl)/(2*(n_cases + n_controls)) 
  
  #calculate per-sd betas
  table$beta = (2 * (table$AF_case - table$AF) * prevalence)/sqrt(2 * table$AF * (1 - table$AF) * prevalence * (1 - prevalence))  
  
  #calculate variant variances
  table$twopq = 2*table$AF * (1 - table$AF) 
  
  #convert betas from per-sd (i.e. sqrt(variance explained)) to per-allele (i.e. in units of phenotype) betas.
  #per-allele are the usual betas reported by an exome wide association study.
  table$beta_perallele = table$beta/sqrt(table$twopq)
  
  #aggregate into gene-level table.
  #position doesn't need to be super precise, as it is only used to order genes for jackknife
  #N = sum of case and control counts
  sumstats = data.frame(gene = table$gene_id,
                        AF = table$AF,
                        beta = table$beta_perallele,
                        gene_position = parse_number(sapply(strsplit(table$locus, split = ":"), function(x) x[[2]])),
                        chromosome =  parse_number(sapply(strsplit(table$locus, split = ":"), function(x) x[[1]])),
                        N = n_cases + n_controls,
                        phenotype_key = "BP")
  
  #we have found that in these smaller sample analyses, there are some genes with
  #large burden scores that are clearly outliers
  #we remove one such gene, TTN
  sumstats <- sumstats[sumstats$gene != "ENSG00000155657",]
  
  
  return(sumstats)
}

bp_sumstats_ptv = wrangle_sumstats(bp_variantlevel_bipex,
                                             14210, #N from Palmer et al, 2022
                                             14422, #N from Palmer et al, 2022
                                             bp_variantlevel_bipex$consequence == "ptv")

bp_sumstats_missenseMPC2 = wrangle_sumstats(bp_variantlevel_bipex[!is.na(bp_variantlevel_bipex$mpc),],
                                             14210,
                                             14422,
                                              bp_variantlevel_bipex$consequence[!is.na(bp_variantlevel_bipex$mpc)] %in% c("damaging_missense", "other_missense") &
                                              bp_variantlevel_bipex$mpc[!is.na(bp_variantlevel_bipex$mpc)] > 2)

bp_sumstats_synonymous = wrangle_sumstats(bp_variantlevel_bipex,
                                             14210,
                                             14422,
                                             bp_variantlevel_bipex$consequence == "synonymous")



```

## Running BHR
```{r run}
#install BHR
#devtools::install_github("ajaynadig/bhr", force = TRUE)

bp_ptv_bhr <- BHR(trait1_sumstats = bp_sumstats_ptv, 
                       annotations = list(baseline_model), #baseline model including constraint annotations
                       num_blocks = 100, #number of blocks for jackknife
                       mode = "univariate") #run in univariate mode to compute burden h2


bp_missenseMPC2_bhr <- BHR(bp_sumstats_missenseMPC2, 
                       annotations = list(baseline_model),
                       num_blocks = 100,
                       mode = "univariate")

bp_synonymous_bhr <- BHR(bp_sumstats_synonymous, 
                       annotations = list(baseline_model),
                       num_blocks = 100,
                       mode = "univariate")

#convert observed scale to liability scale h2

#calculate sample prevalence
bp_sample_prevalence = 14210/(14210 + 14422) 

#this population prevalence estimate is from Ferrari et al, 2016 Bipolar Disorders
population_prevalence_bp = 0.007 


obs2lia_factor <- function(K, P){
  X <- qnorm(K,lower.tail=FALSE)
  z <- (1/sqrt(2*pi))*(exp(-(X**2)/2))
  factor <- (K*(1-K)*K*(1-K))/(P*(1-P)*(z**2))
  return(factor)
}

bp_scalingfactor = obs2lia_factor(population_prevalence_bp,bp_sample_prevalence)

print(paste0("BP PTV Burden Heritability: ",
             round(bp_ptv_bhr$mixed_model$heritabilities[1,5] * bp_scalingfactor, 4),
             " ",
             "(",
             round(bp_ptv_bhr$mixed_model$heritabilities[2,5] * bp_scalingfactor, 4),
             ")"))

print(paste0("BP Missense MPC > 2 Burden Heritability: ",
             round(bp_missenseMPC2_bhr$mixed_model$heritabilities[1,5] * bp_scalingfactor, 4),
             " ",
             "(",
             round(bp_missenseMPC2_bhr$mixed_model$heritabilities[2,5] * bp_scalingfactor, 4),
             ")"))

print(paste0("BP Synonymous Burden Heritability: ",
             round(bp_synonymous_bhr$mixed_model$heritabilities[1,5] * bp_scalingfactor, 4),
             " ",
             "(",
             round(bp_synonymous_bhr$mixed_model$heritabilities[2,5] * bp_scalingfactor, 4),
             ")"))


```
